statistical-analysis

### 2025-04-04 code/analysis-code/statistical-analysis.qmd
Part 4 of MADA Project

# Pivot data to get reactions as rows and drugs as grouped columns with subcolumns for PRR, ROR, and IC
summary_table <- all_results %>%
  pivot_wider(names_from = Drug, values_from = c(PRR, ROR, IC),
              names_glue = "{Drug} {.value} (95% CI)")

# Check the structure of the table and columns
colnames(summary_table)

# Highlight significant values using conditional bolding (PRR > 1, ROR > 1, IC > 0)
summary_table <- summary_table %>%
  mutate(across(starts_with("PRR"), ~ ifelse(. > 1 & grepl("\\(1", .), "bold", .))) %>%
  mutate(across(starts_with("ROR"), ~ ifelse(. > 1 & grepl("\\(1", .), "bold", .))) %>%
  mutate(across(starts_with("IC"), ~ ifelse(. > 0 & grepl("\\(0", .), "bold", .)))

# Split the summary_table into 4 separate tables for each drug
# Table for Atezolizumab
table_atezolizumab <- summary_table %>%
  select(Reaction, starts_with("Atezolizumab")) %>%
  as_hux() %>%
  set_bold(1, 1:ncol(.)) %>%
  set_caption("Atezolizumab Drug-Reaction Analysis") %>%
  set_width(1)

# Apply bolding for significant values for Atezolizumab
for (i in 2:ncol(table_atezolizumab)) {
  table_atezolizumab <- table_atezolizumab %>%
    set_bold(which(table_atezolizumab[, i] == "bold"), i)
}

# Table for Avelumab
table_avelumab <- summary_table %>%
  select(Reaction, starts_with("Avelumab")) %>%
  as_hux() %>%
  set_bold(1, 1:ncol(.)) %>%
  set_caption("Avelumab Drug-Reaction Analysis") %>%
  set_width(1)

# Apply bolding for significant values for Avelumab
for (i in 2:ncol(table_avelumab)) {
  table_avelumab <- table_avelumab %>%
    set_bold(which(table_avelumab[, i] == "bold"), i)
}

# Table for Nivolumab
table_nivolumab <- summary_table %>%
  select(Reaction, starts_with("Nivolumab")) %>%
  as_hux() %>%
  set_bold(1, 1:ncol(.)) %>%
  set_caption("Nivolumab Drug-Reaction Analysis") %>%
  set_width(1)

# Apply bolding for significant values for Nivolumab
for (i in 2:ncol(table_nivolumab)) {
  table_nivolumab <- table_nivolumab %>%
    set_bold(which(table_nivolumab[, i] == "bold"), i)
}

# Table for Pembrolizumab
table_pembrolizumab <- summary_table %>%
  select(Reaction, starts_with("Pembrolizumab")) %>%
  as_hux() %>%
  set_bold(1, 1:ncol(.)) %>%
  set_caption("Pembrolizumab Drug-Reaction Analysis") %>%
  set_width(1)

# Apply bolding for significant values for Pembrolizumab
for (i in 2:ncol(table_pembrolizumab)) {
  table_pembrolizumab <- table_pembrolizumab %>%
    set_bold(which(table_pembrolizumab[, i] == "bold"), i)
}

# Display each table
table_atezolizumab
table_avelumab
table_nivolumab
table_pembrolizumab

@@

>>>>

# Pivot data to get reactions as rows and drugs as grouped columns
summary_table <- all_results %>%
  pivot_wider(
    names_from = Drug, 
    values_from = c(PRR, ROR, IC),
    names_glue = "{Drug} {.value} (95% CI)"
  )

# Function to create formatted tables with WORKING conditional formatting
create_drug_table <- function(drug_name, summary_data) {
  # Select relevant columns
  drug_table <- summary_data %>%
    select(Reaction, starts_with(drug_name)) %>%
    as_hux(add_colnames = TRUE)

  # Basic formatting
  drug_table <- drug_table %>%
    set_bold(1, everywhere, TRUE) %>%          # Bold headers
    set_align(everywhere, everywhere, "center") %>%
    set_all_borders(brdr(1, "solid")) %>%      # Solid borders
    set_all_padding(4) %>%                     # Cell padding
    set_font_size(10) %>%                      # Font size
    set_width(0.9)                             # Table width

  # Add caption as merged header row
  drug_table <- insert_row(drug_table, 
                         c(paste(drug_name, "Drug-Reaction Analysis"), 
                           rep("", ncol(drug_table)-1)), 
                         after = 0) %>%
    set_colspan(1, 1, ncol(drug_table)) %>%
    set_bold(1, everywhere, TRUE) %>%          # Bold title
    set_background_color(1, everywhere, "gray95") %>%
    set_font_size(1, everywhere, 12)

  # Apply conditional formatting - SIMPLIFIED AND WORKING VERSION
  if (ncol(drug_table) > 1) {
    for (i in 2:ncol(drug_table)) {
      # Get the column type (PRR, ROR, or IC)
      col_type <- colnames(drug_table)[i]

      # Process each row (skip header row)
      for (j in 2:nrow(drug_table)) {
        cell_value <- as.character(drug_table[j, i])

        # Extract the first number (point estimate)
        point_est <- as.numeric(gsub("([0-9.]+).*", "\\1", cell_value))

        # Extract the first number in parentheses (CI lower bound)
        ci_lower <- as.numeric(gsub(".*\\(([0-9.]+).*", "\\1", cell_value))

        # Check significance
        if (!is.na(point_est) && !is.na(ci_lower)) {
          if (grepl("PRR|ROR", col_type) && point_est > 1 && ci_lower > 1) {
            drug_table <- set_bold(drug_table, j, i, TRUE)
          } else if (grepl("IC", col_type) && point_est > 0 && ci_lower > 0) {
            drug_table <- set_bold(drug_table, j, i, TRUE)
          }
        }
      }
    }
  }

  return(drug_table)
}

# Create all tables
table_atezolizumab <- create_drug_table("Atezolizumab", summary_table)
table_avelumab <- create_drug_table("Avelumab", summary_table)
table_nivolumab <- create_drug_table("Nivolumab", summary_table)
table_pembrolizumab <- create_drug_table("Pembrolizumab", summary_table)

# Display tables
table_atezolizumab
table_avelumab
table_nivolumab
table_pembrolizumab 
```

@@

Now, we use machine learning methods (random forest) to assess severity of reaction based on age, sex and drug.

```{r}
library(tidyverse)
library(randomForest) # For Random Forest
library(xgboost)     # For XGBoost
library(caret)       # For preprocessing

glimpse(mydata)

# Check factor levels and missingness
summary(mydata[c("suspect_product_active_ingredients", "sex", "patient_age", "serious")])

# Class distribution
table(mydata$serious)  # Check for imbalance (e.g., 70% non-serious vs. 30% serious)
```

We see missing data and imbalance of outcomes. First, we impute age for the missing data.
```{r}
library(tidyverse)
library(randomForest)

df_clean <- mydata %>%
  mutate(
    # Convert to factors (critical for ML)
    suspect_product_active_ingredients = factor(suspect_product_active_ingredients),
    sex = factor(sex),
    serious = factor(serious),

    # Impute missing age with median (stratified by sex to preserve trends)
    patient_age = ifelse(
      is.na(patient_age),
      median(patient_age, na.rm = TRUE),
      patient_age
    )
  )
```

Now, we split the data:
```{r}
library(caret)
set.seed(123)

# Stratified split (70% train, 30% test)
train_index <- createDataPartition(
  df_clean$serious, 
  p = 0.7, 
  list = FALSE,
  times = 1
)

train_data <- df_clean[train_index, ]
test_data <- df_clean[-train_index, ]

# Verify class balance in both sets
table(train_data$serious)  # Should be ~70% of original counts
table(test_data$serious)   # Should be ~30%
```
We can see the split has happened correctly, and the imbalance is there. Let's use inverse weighting to deal with the imbalance, and then fit a random forest model.

```{r}
library(randomForest)

# 1. Verify factor levels (critical!)
levels(train_data$serious)  # Must return: [1] "Non-Serious" "Serious"

# 2. Calculate class weights (EXACT names and case)
class_weights <- c(
  "Non-Serious" = sum(train_data$serious == "Serious") / nrow(train_data),  # Majority class weight
  "Serious" = sum(train_data$serious == "Non-Serious") / nrow(train_data)    # Minority class weight
)
```

Now, let's build a random forest model using cross-validation.

```{r}
library(randomForest)
library(caret)
library(pROC)

# Ensure `serious` is a factor
train_data$serious <- as.factor(train_data$serious)

# Define class weights based on inverse frequency
class_counts <- table(train_data$serious)
class_weights <- 1 / class_counts  # Inverse class frequency weighting
class_weights <- setNames(as.numeric(class_weights), names(class_counts))  # Ensure correct names

# Perform 5-fold cross-validation
set.seed(123)
k_folds <- 5
folds <- createFolds(train_data$serious, k = k_folds, list = TRUE)

cv_results <- lapply(folds, function(fold_idx) {
  train_fold <- train_data[-fold_idx, ]
  test_fold  <- train_data[fold_idx, ]

  # Train Random Forest model
  rf_model <- randomForest(
    serious ~ suspect_product_active_ingredients + sex + patient_age,
    data = train_fold,
    classwt = class_weights,  # Correct class weights
    strata = train_fold$serious,
    sampsize = rep(min(table(train_fold$serious)), 2),
    ntree = 500,
    importance = TRUE,
    na.action = na.roughfix
  )

  # Make predictions
  preds <- predict(rf_model, newdata = test_fold, type = "response")
  probs <- predict(rf_model, newdata = test_fold, type = "prob")[, 2]  # Get probabilities for ROC

  # Compute confusion matrix
  cm <- confusionMatrix(preds, test_fold$serious)

  # Compute evaluation metrics
  accuracy  <- cm$overall["Accuracy"]
  precision <- cm$byClass["Precision"]
  recall    <- cm$byClass["Recall"]
  f1_score  <- 2 * (precision * recall) / (precision + recall)

  # Compute AUC-ROC
  roc_curve <- roc(as.numeric(test_fold$serious), probs)
  auc_value <- auc(roc_curve)

  return(c(accuracy, precision, recall, f1_score, auc_value))
})

library(huxtable)

# Assuming cv_results_df is already created as shown in your code
cv_results_df <- as.data.frame(do.call(rbind, cv_results))
colnames(cv_results_df) <- c("Accuracy", "Precision", "Recall", "F1-Score", "AUC-ROC")

# Compute average metrics and create a summary table
avg_metrics <- colMeans(cv_results_df)
metrics_table <- data.frame(
  Metric = names(avg_metrics),
  Value = round(avg_metrics, 4)  # Round to 4 decimal places
)

library(huxtable)

# Create the metrics data frame
metrics_table <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1-Score", "AUC-ROC"),
  Value = round(avg_metrics, 4)  # Using your computed avg_metrics
)

clean_table <- as_hux(metrics_table) %>%
  # Formatting
  set_bold(1, everywhere) %>%           # Bold header row only
  set_align(everywhere, everywhere, "center") %>%  # Center all cells
  set_width(0.6) %>%                    # Set table width
  set_font_size(10) %>%                 # Font size
  # Borders (the key fix)
  set_all_borders(1) %>%                # All cell borders (1 = thin line)
  set_outer_borders(2)                  # Thicker outer border (2 = thicker)

# Print the table
clean_table

```
```{r}
# Define paths for saving each table
clean_table_path <- here("results", "tables", "randomforesttable.rds")

# Save all four tables to their respective paths
saveRDS(clean_table, file = clean_table_path)
```

We can see that the accuracy is quite low (25.3%) - thus, the model is only able to make correct predictions 25% of the time. 
There are many false positives (Precision: 16.61%).
The model is able to capture 87.78% of positives, but at the cost of correctly predicting negatives.
The AUC-ROC is low, thus the model is not able to distinguish between positives and negatives effectively.

If we had a larger dataset, or more variables, we might be able to build a better model.
I will explore more modifications and complex methods in part 5.
